Missing the Point: Non-Convergence in Iterative Imputation Algorithms

### 2020-05-11 3.Thesis/Writeup/2.Manuscript.Rmd
Update reference list

By visually inspecting the imputation chains in panel A, we conclude that the two scenarios show very different convergence. The typically converged algorithm initially portrayes some signs of non-mixing (around $t=2$), but intermingles nicely overall. Additionally, there is very little trending in this scenario. The algorithm with pathological non-convergence shows severe non-mixing, although this gradually improves (beyond $t=7$). In this scenario, we see a lot of trending initially (up-to $t=6$), after which the chains reached a somewhat more stationary state.

@@

2. Our results also support the hyothesis that $\widehat{R}$ and $AC$ would correctly identify signs of non-convergence. We observed higher $\widehat{R}$ and $AC$-values in conditions where $\bar{Q}$s were biased, invalid estimates of $Q$s.

@@

Much remains unknown about non-convergence in iterative imputation algorithms. Even though we have demonstrated the appropriateness of $\widehat{R}$ and $AC$ as non-convergence diagnostics in this study, results may not extrapolate to other situations. The current simulation conditions were restricted to a single missingness mechanism. Proper performance under a 'missing completely at random' (MCAR) mechanism is a necessary condition for any missing data method. It does not't, however, guarantee equal performance under different missingness mechanisms. Future research should determine the performance of $\widehat{R}$ and $AC$ under 'missing at random' and 'missing not at random' mechanisms. 
Another parameter to consider in future research is the choice of the imputation method. As @murr18 concluded, the behavior of algorithms such as MICE under certain default imputation models is still an open research question. We have investigated the behavior of MICE under only one type of model (Bayesian linear regression). Different imputations models might converge more poorly. Moreover, the imputation model may even be mis-specified on purpose to induce different levels of non-convergence. Qualitatively, we have shown that $\widehat{R}$ and $AC$ are appropriate under clear violation of convergence. Quantitatively, we have only demonstrated that they can identify signs of non-convergence under MCAR, when the imputation model is a correctly specified Bayesian linear regression model. **[ADD limitation that other MI packages may have different way to obtain initial values!]** \newline \newline

>>>>

By visually inspecting the imputation chains in panel A, we conclude that the two scenarios show very different convergence. The typically converged algorithm initially portrays some signs of non-mixing (around $t=2$), but intermingles nicely overall. Additionally, there is very little trending in this scenario. The algorithm with pathological non-convergence shows severe non-mixing, although this gradually improves (beyond $t=7$). In this scenario, we see a lot of trending initially (up-to $t=6$), after which the chains reached a somewhat more stationary state.

@@

2. Our results also support the hypothesis that $\widehat{R}$ and $AC$ would correctly identify signs of non-convergence. We observed higher $\widehat{R}$ and $AC$-values in conditions where $\bar{Q}$s were biased, invalid estimates of $Q$s.

@@

Much remains unknown about non-convergence in iterative imputation algorithms. Even though we have demonstrated the appropriateness of $\widehat{R}$ and $AC$ as non-convergence diagnostics in this study, results may not extrapolate to other situations. The current simulation conditions were restricted to a single missingness mechanism. Proper performance under a 'missing completely at random' (MCAR) mechanism is a necessary condition for any missing data method. It does not, however, guarantee equal performance under different missingness mechanisms. Future research should determine the performance of $\widehat{R}$ and $AC$ under 'missing at random' and 'missing not at random' mechanisms. 
Another parameter to consider in future research is the choice of the imputation method. As @murr18 concluded, the behavior of algorithms such as MICE under certain default imputation models is still an open research question. We have investigated the behavior of MICE under only one type of model (Bayesian linear regression). Different imputations models might converge more poorly. Moreover, the imputation model may even be mis-specified on purpose to induce different levels of non-convergence. Qualitatively, we have shown that $\widehat{R}$ and $AC$ are appropriate under clear violation of convergence. Quantitatively, we have only demonstrated that they can identify signs of non-convergence under MCAR, when the imputation model is a correctly specified Bayesian linear regression model. **[ADD limitation that other MI packages may have different way to obtain initial values!]** \newline \newline

### 2020-05-11 3.Thesis/Writeup/2.Manuscript.tex
Update reference list

By visually inspecting the imputation chains in panel A, we conclude that the two scenarios show very different convergence. The typically converged algorithm initially portrayes some signs of non-mixing (around \(t=2\)), but intermingles nicely overall. Additionally, there is very little trending in this scenario. The algorithm with pathological non-convergence shows severe non-mixing, although this gradually improves (beyond \(t=7\)). In this scenario, we see a lot of trending initially (up-to \(t=6\)), after which the chains reached a somewhat more stationary state.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  As hypothesized, biased, invalid estimates of quantities of scientific interest occurred more often in simulation conditions with a high proportion of incomplete cases \(p_{\rm inc}\) and a low number of iterations \(T\). However, this only holds for multivariate scientific estimands. Univariate \(Q\)s are not substantially impacted by early stopping.
\item
  Our results also support the hyothesis that \(\widehat{R}\) and \(AC\) would correctly identify signs of non-convergence. We observed higher \(\widehat{R}\) and \(AC\)-values in conditions where \(\bar{Q}\)s were biased, invalid estimates of \(Q\)s.
\item
  We hypothesized that the recommended thresholds to diagnose non-convergence with \(\widehat{R}\) would be too stringent for iterative imputation applications. This hypothesis is only partially supported. \(\widehat{R}\) indicated improving convergence until 30 to 40 iterations, while approximately unbiased, confidence-valid estimates were obtained after just seven iterations. This roughly corresponds to the threshold of 1.2. The more stringent threshold 1.1 seems plausible compared to the decrease in \(\widehat{R}\)-values across iterations. The most recent threshold 1.01 does indeed seem to over-estimate the signs of non-convergence.
\item
  The results of this simulation study do not support the hypothesis that high \(AC\)-values are implausible in iterative imputation algorithms with typical convergence. We expected that the randomness induced by the algorithm would effectively mitigate the risk of dependency within chains after a few iterations. In this study, however, we observed \(AC\)-values as high as theoretically possible (\(AC=1\)). The second part of the hypothesis is not refuted: the \(AC\)-values did decrease quickly as a function of \(T\).
\item
  Our final hypothesis was that multivariate \(\theta\)s would be better at detecting non-convergence than univariate \(\theta\)s. Multivariate \(\theta\)s did indeed show superior performance under certain conditions (e.g., multivariate \(Q\)s and extreme missingness proportions).
\end{enumerate}

@@

Much remains unknown about non-convergence in iterative imputation algorithms. Even though we have demonstrated the appropriateness of \(\widehat{R}\) and \(AC\) as non-convergence diagnostics in this study, results may not extrapolate to other situations. The current simulation conditions were restricted to a single missingness mechanism. Proper performance under a `missing completely at random' (MCAR) mechanism is a necessary condition for any missing data method. It does not't, however, guarantee equal performance under different missingness mechanisms. Future research should determine the performance of \(\widehat{R}\) and \(AC\) under `missing at random' and `missing not at random' mechanisms.
Another parameter to consider in future research is the choice of the imputation method. As \citet{murr18} concluded, the behavior of algorithms such as MICE under certain default imputation models is still an open research question. We have investigated the behavior of MICE under only one type of model (Bayesian linear regression). Different imputations models might converge more poorly. Moreover, the imputation model may even be mis-specified on purpose to induce different levels of non-convergence. Qualitatively, we have shown that \(\widehat{R}\) and \(AC\) are appropriate under clear violation of convergence. Quantitatively, we have only demonstrated that they can identify signs of non-convergence under MCAR, when the imputation model is a correctly specified Bayesian linear regression model. \textbf{{[}ADD limitation that other MI packages may have different way to obtain initial values!{]}} \newline \newline

>>>>

By visually inspecting the imputation chains in panel A, we conclude that the two scenarios show very different convergence. The typically converged algorithm initially portrays some signs of non-mixing (around \(t=2\)), but intermingles nicely overall. Additionally, there is very little trending in this scenario. The algorithm with pathological non-convergence shows severe non-mixing, although this gradually improves (beyond \(t=7\)). In this scenario, we see a lot of trending initially (up-to \(t=6\)), after which the chains reached a somewhat more stationary state.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  As hypothesized, biased, invalid estimates of quantities of scientific interest occurred more often in simulation conditions with a high proportion of incomplete cases \(p_{\rm inc}\) and a low number of iterations \(T\). However, this only holds for multivariate scientific estimands. Univariate \(Q\)s are not substantially impacted by early stopping.
\item
  Our results also support the hypothesis that \(\widehat{R}\) and \(AC\) would correctly identify signs of non-convergence. We observed higher \(\widehat{R}\) and \(AC\)-values in conditions where \(\bar{Q}\)s were biased, invalid estimates of \(Q\)s.
\item
  We hypothesized that the recommended thresholds to diagnose non-convergence with \(\widehat{R}\) would be too stringent for iterative imputation applications. This hypothesis is only partially supported. \(\widehat{R}\) indicated improving convergence until 30 to 40 iterations, while approximately unbiased, confidence-valid estimates were obtained after just seven iterations. This roughly corresponds to the threshold of 1.2. The more stringent threshold 1.1 seems plausible compared to the decrease in \(\widehat{R}\)-values across iterations. The most recent threshold 1.01 does indeed seem to over-estimate the signs of non-convergence.
\item
  The results of this simulation study do not support the hypothesis that high \(AC\)-values are implausible in iterative imputation algorithms with typical convergence. We expected that the randomness induced by the algorithm would effectively mitigate the risk of dependency within chains after a few iterations. In this study, however, we observed \(AC\)-values as high as theoretically possible (\(AC=1\)). The second part of the hypothesis is not refuted: the \(AC\)-values did decrease quickly as a function of \(T\).
\item
  Our final hypothesis was that multivariate \(\theta\)s would be better at detecting non-convergence than univariate \(\theta\)s. Multivariate \(\theta\)s did indeed show superior performance under certain conditions (e.g., multivariate \(Q\)s and extreme missingness proportions).
\end{enumerate}

@@

Much remains unknown about non-convergence in iterative imputation algorithms. Even though we have demonstrated the appropriateness of \(\widehat{R}\) and \(AC\) as non-convergence diagnostics in this study, results may not extrapolate to other situations. The current simulation conditions were restricted to a single missingness mechanism. Proper performance under a `missing completely at random' (MCAR) mechanism is a necessary condition for any missing data method. It does not, however, guarantee equal performance under different missingness mechanisms. Future research should determine the performance of \(\widehat{R}\) and \(AC\) under `missing at random' and `missing not at random' mechanisms.
Another parameter to consider in future research is the choice of the imputation method. As \citet{murr18} concluded, the behavior of algorithms such as MICE under certain default imputation models is still an open research question. We have investigated the behavior of MICE under only one type of model (Bayesian linear regression). Different imputations models might converge more poorly. Moreover, the imputation model may even be mis-specified on purpose to induce different levels of non-convergence. Qualitatively, we have shown that \(\widehat{R}\) and \(AC\) are appropriate under clear violation of convergence. Quantitatively, we have only demonstrated that they can identify signs of non-convergence under MCAR, when the imputation model is a correctly specified Bayesian linear regression model. \textbf{{[}ADD limitation that other MI packages may have different way to obtain initial values!{]}} \newline \newline
